---
title: mQTL p-values by type of association
output: html_document
---

I was a bit concerned that the distribution of pvalues looked bad - very many p-values of 0. I just wanted to see what what the random effects p-values looked like by comparison.

```{r}
library(knitr)
library(dplyr)
library(ggplot2)
library(tidyr)

load("../../results/16/16_clumped.rdata")
```

How many p-values are 0 for different models?

```{r}
group_by(clumped, cis == "TRUE") %>%
	summarise(
		count=n(), 
		prop_fixed_0=sum(pval == 0) / n(),
		prop_ARE_0=sum(PvalueARE == 0) / n(),
		prop_MRE_0=sum(PvalueMRE == 0) / n()
	) %>%
	kable
```

Lots of pvalues of 0 for fixed effects model. They seem to be typically those with a large number of effects in the same direction

```{r}
table(subset(clumped, pval == 0)$Direction) %>% sort(decr=TRUE) %>% as.data.frame %>% head(30) %>% kable
```

What is the distribution of p-values for different meta analysis models? Fix 0 to the smallest non-0 p-value:

```{r}
minval <- subset(clumped, pval != 0)$pval %>% min
clumped$pval[clumped$pval == 0] <- minval
clumped$PvalueARE[clumped$PvalueARE == 0] <- minval
clumped$PvalueMRE[clumped$PvalueMRE == 0] <- minval

temp <- gather(subset(clumped, select=c(pval, PvalueMRE, PvalueARE)))
ggplot(temp, aes(x=-log10(value))) +
geom_histogram() +
facet_grid(key ~ .)
```

What is the distribution for cis-only, trans-only and ambivalent CpGs and SNPs? Use threshold of `cis = 1e-10` and `trans = 1e-14` to calculate this (it will vary for different thresholds)

```{r}

sig <- subset(clumped, (pval < 1e-10 & cis) | (pval < 1e-14 & !cis) )
snptype <- group_by(sig, snp) %>%
	summarise(snpt=sum(unique(as.numeric(cis) + 1)))
snptype$snpt <- dplyr::recode(as.character(snptype$snpt), "1"="Trans only", "2"="Cis only", "3"="Ambivalent")

cpgtype <- group_by(sig, cpg) %>%
	summarise(cpgt=sum(unique(as.numeric(cis) + 1)))
cpgtype$cpgt <- dplyr::recode(as.character(cpgtype$cpgt), "1"="Trans only", "2"="Cis only", "3"="Ambivalent")

sig <- merge(sig, snptype, by="snp")
sig <- merge(sig, cpgtype, by="cpg")

```

Count the cis/trans/ambivalent SNPs and CpGs

```{r}
table(cpgtype$cpgt)
```

```{r}
table(snptype$snpt)
```

**The ambivalent SNPs are the most uncommon - is this correct? When we were discussing I think it was the trans only that were problematic?**

What proportion of associations are significant for ARE and MRE, grouped by type?

```{r}
sig$sigARE <- (sig$PvalueARE < 1e-10 & sig$cis) | (sig$PvalueARE < 1e-14 & !sig$cis)
sig$sigMRE <- (sig$PvalueMRE < 1e-10 & sig$cis) | (sig$PvalueMRE < 1e-14 & !sig$cis)

group_by(sig, snpt, cpgt) %>%
summarise(psigARE = sum(sigARE)/n(), psigMRE=sum(sigMRE)/n(), count=n()) %>%
kable

group_by(sig, snpt) %>%
summarise(psigARE = sum(sigARE)/n(), psigMRE=sum(sigMRE)/n(), count=n()) %>%
kable

group_by(sig, cpgt, cpgt) %>%
summarise(psigARE = sum(sigARE)/n(), psigMRE=sum(sigMRE)/n(), count=n()) %>%
kable
```

Distribution of pvalues:

```{r}
gather(subset(sig, select=c(cpgt, pval, PvalueMRE, PvalueARE)), key="key", value="value", pval, PvalueMRE, PvalueARE) %>%
ggplot(aes(x=-log10(value))) +
geom_histogram(aes(fill=cpgt)) +
facet_grid(cpgt ~ key, scale="free_y") +
scale_fill_brewer(type="qual")

gather(subset(sig, select=c(snpt, pval, PvalueMRE, PvalueARE)), key="key", value="value", pval, PvalueMRE, PvalueARE) %>%
ggplot(aes(x=-log10(value))) +
geom_histogram(aes(fill=snpt)) +
facet_grid(snpt ~ key, scale="free_y") +
scale_fill_brewer(type="qual")
```

The distributions of p-values for the different models are extremely similar, suggesting that there is nothing systematically wrong with the results for cis-only, trans-only or ambivalent associations.
