---
title: Are intra and inter chromosomal trans hits occurring at the same rate?
output:
  html_document:
    toc: true
    theme: united
---

```{r}
suppressWarnings(suppressPackageStartupMessages({
	library(knitr)
	library(dplyr)
}))
opts_chunk$set(cache=TRUE, echo=TRUE, message=FALSE, warning=FALSE)
```

We have a 1mb radius defining cis. If a trans hit is 1-6mb away, is that as common as something 6-11mb, or in any 5mb region on another chromosome?

Only analyse significant cis and trans hits, and remove chr 6 for simplicity

```{r}
load("../results/16/16_clumped.rdata")
sig <- subset(clumped, 
	(cis & pval < 1e-8) | (!cis & pval < 1e-14)
)
sig <- subset(sig, cpgchr != "chr6" & snpchr != "chr6")
index <- sig$cpgchr == sig$snpchr
sig$cisdist <- NA
sig$cisdist[index] <- abs(sig$cpgpos - sig$snppos)[index]
```

Distribution of distance from cpg for cis

```{r}
hist(sig$cisdist[sig$cis], breaks=100)
```

Distribution in 5 mb region

```{r}
hist(sig$cisdist[sig$cisdist < 5000000], breaks=100)
```


Question: is the 1mb - 6mb region containing more trans than 6mb-11mb

```{r}
sum(!is.na(sig$cisdist[sig$cisdist > 1000000 & sig$cisdist <= 6000000]))
sum(!is.na(sig$cisdist[sig$cisdist > 6000000 & sig$cisdist <= 11000000]))
```

How often are CpGs close to the chromosome boundary? Note that there will be more CpGs unable to have a further trans effect because of chromosome boundaries

```{r}
chromsizes <- read.table("chromosome_sizes.txt", header=TRUE, stringsAsFactors=FALSE)
chromsizes$chrsize / 1000000

# What is the size of the inter chromosomal region for each mQTL?
chromsizes$inter_size <- sum(chromsizes$chrsize) - chromsizes$chrsize

sig <- merge(sig, chromsizes, by.x="cpgchr", by.y="chr")
sapply(c(1:11) * 1000000, function(x)
{
	sum((sig$cpgpos - x) < 0 | (sig$cpgpos + x) > sig$chrsize ) / nrow(sig)
})
```

Restrict only to CpGs that are within the relevant window

```{r}
# sum((sig$cpgpos - 6000000) > 0 & (sig$cpgpos + 6000000) < sig$chrsize) 
# sum((sig$cpgpos - 11000000) > 0 & (sig$cpgpos + 11000000) < sig$chrsize)

sum(
	(sig$cpgpos - 6000000) >= 0 & 
	(sig$cpgpos + 6000000) <= sig$chrsize &
	(sig$cisdist > 1000000 & sig$cisdist <= 6000000),
	na.rm=TRUE
)

sum(
	(sig$cpgpos - 11000000) >= 0 & 
	(sig$cpgpos + 11000000) <= sig$chrsize &
	(sig$cisdist > 6000000 & sig$cisdist <= 11000000),
	na.rm=TRUE
)
```


Question: is the intra-chromosomal trans region containing more trans hits than inter chromosomal regions.

```{r}
p1 <- subset(sig, cpgchr == snpchr & abs(cpgpos - snppos) > 1000000) %>%
	group_by(cpgchr) %>%
	summarise(
		n=n(), 
		intra_size=first(chrsize),
		prop=n/(intra_size / 5000000)
	) %>%
	summarise(m= sum(prop*n) / sum(n)) %>%
	unlist()

p2 <- subset(sig, cpgchr == snpchr & abs(cpgpos - snppos) > 6000000) %>%
	group_by(cpgchr) %>%
	summarise(
		n=n(), 
		intra_size=first(chrsize),
		prop=n/(intra_size / 5000000)
	) %>%
	summarise(m= sum(prop*n) / sum(n)) %>%
	unlist()

p3 <- subset(sig, cpgchr != snpchr) %>%
	group_by(cpgchr) %>%
	summarise(
		n=n(), 
		inter_size=first(inter_size),
		prop=n/(inter_size / 5000000)
	) %>%
	summarise(m= sum(prop*n) / sum(n)) %>%
	unlist()
```

Number of trans mQTLs per 5mb region when:

- More than 1mb away on the same chromosome: `r p1`
- More than 6mb away on the same chromosome: `r p2`
- On a different chromosome: `r p3`
