---
title: Following up colocalisation signals with secondary mQTLs (analysis)
author: Gibran Hemani
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

```{r}
suppressWarnings(suppressPackageStartupMessages({
	library(knitr)
	library(dplyr)
	library(TwoSampleMR)
	library(ggplot2)
	library(ggthemes)
	library(magrittr)
	library(ggrepel)
}))
opts_chunk$set(cache=TRUE, echo=TRUE, message=FALSE, warning=FALSE)
```

```{r}
threshold1 <- 0.05 / (300000 * 698)
threshold2 <- 0.05 / (300000)
load("../results/mr_disc_repl.rdata")
load("../../05_cis-trans-networks/data/outcomes.rdata")
ao <- ao %>% filter(access != "developer")
```

```{r}
ggplot(mr_disc_repl %>% filter(!is.na(z.repl)), aes(y=z.repl, x=z.disc)) +
geom_point(aes(colour=fdr < 0.05)) +
geom_abline(slope=1, linetype="dotted") +
geom_smooth(method="lm") +
ylim(min(mr_disc_repl$z.disc), max(mr_disc_repl$z.disc)) +
labs(x="Z-score primary MR", y="Z-score secondary MR", colour="Replication\nFDR < 0.05")
```

```{r}
sub <- subset(mr_disc_repl, abs(b.repl) > 1.5 & abs(b.repl) < 20 | (abs(b.disc) > 10 & b.repl < 0) & b.repl > -50 ) %>% 
	tidyr::separate(outcome.disc, sep=" \\|", into=c("label", "other"))
ggplot(mr_disc_repl %>% filter(!is.na(z.repl), b.repl > -50), aes(y=b.repl, x=b.disc)) +
geom_point(aes(colour=fdr < 0.05)) +
geom_abline(slope=1, linetype="dotted") +
geom_smooth(method="lm") +
geom_text_repel(data=sub, aes(label=label)) +
labs(x="Z-score primary MR", y="Z-score secondary MR", colour="Replication\nFDR < 0.05")
```


```{r}
summary(lm(b.repl ~ b.disc,  mr_disc_repl))
summary(lm(b.repl ~ b.disc,  mr_disc_repl %>% filter(b.repl > -50)))
```

Are there more small p-values than expected?

```{r}
binom.test(x=sum(mr_disc_repl$pval.repl < 0.05, na.rm=T), n=nrow(mr_disc_repl), p=0.05)
```

What about direction of effect?

```{r}
binom.test(x=sum(sign(mr_disc_repl$b.disc) == sign(mr_disc_repl$b.repl)), n=nrow(mr_disc_repl), p=0.5)
```

What about direction of effect for just significant assocs?

```{r}
subset(mr_disc_repl, pval.repl < 0.05) %$% binom.test(x=sum(sign(b.disc) == sign(b.repl)), n=length(pval.repl), p=0.5)
```

Is small p-values in repl related to trait type

```{r}
inner_join(mr_disc_repl, subset(ao, select=c(id, subcategory)), by=c("id.outcome"="id")) %$% summary(lm(z.repl ~ subcategory))
```

```{r}
inner_join(mr_disc_repl, subset(ao, select=c(id, subcategory)), by=c("id.outcome"="id")) %>% mutate(sig=pval.repl < 0.05) %$% summary(glm(sig ~ subcategory))
```

```{r}
inner_join(mr_disc_repl, subset(ao, select=c(id, subcategory)), by=c("id.outcome"="id")) %>% mutate(sig=pval.repl < 0.05, blood = subcategory %in% c("Haemotological", "Metal", "Protein")) %$% summary(glm(sig ~ blood))
```

Which associations appear to be working?

```{r}

mr_disc_repl$fdr <- p.adjust(mr_disc_repl$pval.repl, "fdr")
mr_disc_repl$concordant <- sign(mr_disc_repl$b.repl) == sign(mr_disc_repl$b.disc)
subset(mr_disc_repl, fdr < 0.05 & concordant)
subset(mr_disc_repl, fdr < 0.05 & !concordant)


```

Schizophrenia has some consistent results but looking at all of them together is less convincing

```{r}
ggplot(mr_disc_repl %>% filter(id.outcome == 22), aes(y=z.repl, x=z.disc)) +
geom_point(aes(colour=pval.repl < 0.05)) +
geom_abline(slope=1, linetype="dotted") +
geom_smooth(method="lm") +
ylim(min(mr_disc_repl$z.disc), max(mr_disc_repl$z.disc)) +
labs(x="Z-score primary MR", y="Z-score secondary MR", colour="Replication\np-value < 0.05")

```

LDL cholesterol

```{r}

ggplot(mr_disc_repl %>% filter(id.outcome == 300), aes(y=b.repl, x=b.disc)) +
geom_point(aes(colour=pval.repl < 0.05)) +
geom_abline(slope=1, linetype="dotted") +
geom_smooth(method="lm")
# ylim(min(mr_disc_repl$z.disc), max(mr_disc_repl$z.disc)) +
labs(x="Z-score primary MR", y="Z-score secondary MR", colour="Replication\np-value < 0.05")
```

Systemic lupus erythematosus

```{r}
ggplot(mr_disc_repl %>% filter(id.outcome == 288), aes(y=b.repl, x=b.disc)) +
geom_point(aes(colour=pval.repl < 0.05)) +
geom_abline(slope=1, linetype="dotted") +
geom_smooth(method="lm")
# ylim(min(mr_disc_repl$z.disc), max(mr_disc_repl$z.disc)) +
labs(x="Z-score primary MR", y="Z-score secondary MR", colour="Replication\np-value < 0.05")

```

Create table

```{r}
out <- mr_disc_repl %$% data_frame(
	cpg = exposure,
	outcome = outcome.disc,
	beta.primary = b.disc,
	se.primary = se.disc,
	pval.primary = pval.disc,
	beta.secondary = b.repl,
	se.secondary = se.repl,
	pval.secondary = pval.repl
)

write.csv(out, file="../results/mr_primary_secondary.csv")

```
	

