---
title: MR and colocalisation of DNAm on tissue counts
author: Gibran Hemani
output:
  html_document:
    toc: true
    theme: united
---

```{r}
suppressWarnings(suppressPackageStartupMessages({
	library(dplyr)
	library(magrittr)
	library(data.table)
	library(ggplot2)
	library(tidyr)
	library(knitr)
	library(pander)
}))
opts_chunk$set(cache=TRUE, echo=TRUE, message=FALSE, warning=FALSE, dev="pdf")
```

Read in data

```{r}
load("results/cellcount_coloc.rdata")
load("results/cellcount_mr.rdata")
mrivw$fdr <- p.adjust(mrivw$pval, "fdr")
wr$fdr <- p.adjust(wr$pval, "fdr")
```

Significant ivw

```{r}
mrivw_sig <- subset(mrivw, fdr < 0.05)
nrow(mrivw_sig) / nrow(mrivw)
```

Significant hits more likely when number of instruments is larger

```{r}
table(mrivw_sig$nsnp) / table(mrivw$nsnp)
summary(glm(I(fdr < 0.05) ~ nsnp, mrivw, family="binomial")) %>% pander::pander()
```


Significant wr

```{r}
wr_sig <- subset(wr, fdr < 0.05 & nsnp == 1)
table(wr_sig$method) %>% pander::pander()
table(wr_sig$nsnp) %>% pander::pander()
table(wr$nsnp) %>% pander::pander()
```


Check that the WR and coloc analyses have the same data

```{r}
res$id.exposure <- paste(res$variant, res$cpg)
rescheck <- subset(res, paste(id.exposure, trait) %in% paste(wr_sig$id.exposure, wr_sig$id.outcome))
dim(rescheck) %>% pander::pander()
dim(wr_sig) %>% pander::pander()
```

Coloc

```{r}
table(rescheck$nsnps > 10) %>% pander::pander()
```

```{r}
table(rescheck$PP.H4.abf > 0.8 & rescheck$nsnps > 10) %>% pander::pander()
```

## Comparing colocalising associations against secondary mQTLs

Find colocalising hits

```{r}
wr <- wr %>% tidyr::separate(exposure, sep=" ", into=c("variant", "cpg"), remove=FALSE)

# Get significant wr
wr_sig <- subset(wr, fdr < 0.05)

# Which colocalise?
wr_coloc <- inner_join(
	wr_sig,
	subset(res, PP.H4.abf > 0.8 & nsnps > 10),
	by=c("id.exposure", "outcome"="trait")
)
```

Which of these cpgs have a secondary mqtl?

```{r}
cpg_trait <- unique(paste(wr_coloc$cpg.x, wr_coloc$outcome))
wr_cpg_trait <- subset(wr, paste(cpg, outcome) %in% cpg_trait)
wr_cpg_trait$cpgtrait <- paste(wr_cpg_trait$cpg, wr_cpg_trait$outcome)
table(table(wr_cpg_trait$cpgtrait)) %>% pander::pander()
```

Separate the primary and separate coloc signals

```{r}
wr_cpg_trait_primary <- subset(wr_cpg_trait, paste(id.exposure, outcome) %in% paste(wr_coloc$id.exposure, wr_coloc$id.outcome))

wr_cpg_trait_secondary <- subset(wr_cpg_trait, !paste(id.exposure, outcome) %in% paste(wr_coloc$id.exposure, wr_coloc$id.outcome))
```

For every cpg-trait pair, meta analyse the results to get a single assoc. Essentially creating the IVW fixed effects estimate from the Wald ratios

```{r}
cpg_trait_list <- unique(paste(wr_cpg_trait$cpg, wr_cpg_trait$outcome))
metafn <- function(x)
{
	m <- meta::metagen(x$b, x$se)
	y <- tibble(
		nsnp=nrow(x),
		b=m$TE.fixed,
		se=m$seTE.fixed,
		pval=m$pval.fixed
	)
	y
}
out <- wr_cpg_trait_secondary %>% 
	group_by(cpgtrait) %>%
	do(metafn(.))

```

Combine primary and secondary results

```{r}
ps <- inner_join(wr_cpg_trait_primary, out, by="cpgtrait")
ps$fdr.y <- p.adjust(ps$pval.y, "fdr")
ps$fdr.x <- p.adjust(ps$pval.x, "fdr")
save(ps, file="results/primary-secondary.rdata")
```

Check all primary associations have fdr < 0.05

```{r}
table(ps$fdr.x < 0.05) %>% pander::pander()
```

How many secondary associations have fdr < 0.05?

```{r}
table(ps$fdr.y < 0.05) %>% pander::pander()
```

We should remove chromosome 6 to be careful about MHC region

```{r}
table(ps$fdr.y[!grepl("6:", ps$variant)] < 0.05) %>% pander::pander()
```

Looks like a lot of this is from the MHC region.

Are the primary and secondary effects correlated?

```{r}
summary(lm(b.y ~ b.x, ps)) %>% pander::pander()
```

This effect attenuates sharply after removing MHC

```{r}
summary(lm(b.y ~ b.x, ps %>% subset(., !grepl("6:", variant)))) %>% pander::pander()
```

Look at the plots

```{r}
ggplot(ps %>% subset(., !grepl("6:", variant)), aes(x = b.x, y = b.y)) +
geom_point(aes(colour=fdr.y < 0.05)) +
geom_smooth(method="lm") +
labs(x="Primary causal estimate", y="Secondary causal estimate", colour="Secondary\nFDR < 0.05")
```

A lot of significant results secondary effects, but not much of a relationship between primary and secondary hits.

How does it compare by splitting by number of instruments used for secondary association?

```{r}
ggplot(ps %>% subset(., !grepl("6:", variant)), aes(x = b.x, y = b.y)) +
geom_point(aes(colour=fdr.y < 0.05)) +
geom_smooth(method="lm") +
facet_wrap(~ nsnp.y) +
labs(x="Primary causal estimate", y="Secondary causal estimate", colour="Secondary\nFDR < 0.05")
```

What are the primary-secondary associations?

```{r}
assocfn <- function(x)
{
	mod <- summary(lm(b.y ~ b.x, x))$coefficients
	data.frame(nassoc=nrow(x), b=mod[2,1], se=mod[2,2], p=mod[2,4])	
}

ps %>% 
	subset(., !grepl("6:", variant)) %>%
	group_by(nsnp=nsnp.y) %>%
	do(assocfn(.)) %>%
	pander()
```

Group by blood cell trait category

```{r}
load("bc_type/bc_type.rdata")
bcm$id <- gsub("ebi-a-", "", bcm$id)
stopifnot(all(ps$id.outcome %in% bcm$id))
ps <- inner_join(ps, subset(bcm, select=c(trait, id, type)), by=c("id.outcome"="id"))

ggplot(ps %>% subset(., !grepl("6:", variant) & fdr.y < 0.05), aes(x = b.x, y = b.y)) +
geom_point(aes(colour=type)) +
geom_smooth(method="lm", aes(color=type)) +
labs(x="Primary causal estimate", y="Secondary causal estimate", colour="Secondary\nFDR < 0.05") +
scale_colour_brewer(type="qual", palette=3)


ggplot(ps %>% subset(., !grepl("6:", variant)), aes(x = b.x, y = b.y)) +
geom_point(aes(colour=type)) +
geom_smooth(method="lm", aes(color=type)) +
labs(x="Primary causal estimate", y="Secondary causal estimate", colour="Secondary\nFDR < 0.05") +
scale_colour_brewer(type="qual", palette=3)
```

Coefficients per type

```{r}
ps %>% 
	subset(., !grepl("6:", variant) & fdr.y < 0.05) %>%
	group_by(type=type) %>%
	do(assocfn(.)) %>%
	pander()
ps %>% 
	subset(., !grepl("6:", variant)) %>%
	group_by(type=type) %>%
	do(assocfn(.)) %>%
	pander()

```

Coefficients per study

```{r}
ps %>% 
	subset(., !grepl("6:", variant)) %>%
	group_by(type=trait) %>%
	do(assocfn(.)) %>%
	arrange(desc(b)) %>%
	pander()
```

