---
title: MR of traits on DNAm sites
author: Gibran Hemani
date: 27/02/2019
---

```{r}
suppressWarnings(suppressPackageStartupMessages({
	library(knitr)
	library(TwoSampleMR)
	library(tidyverse)
}))
opts_chunk$set(cache=TRUE, echo=TRUE, message=FALSE, warning=FALSE)
```

Systematically filter out results that look unreliable. Codes are as follows:

```
(1) Fail: Non-significant wald ratio
(2) Pass: Significant wald ratio
(3) Fail: ivw not sig
(4) Pass: 2-9 inst, no heterogeneity
(5) Fail: heterogeneity
(6) Pass: 10-49 inst, no het or high sign
(7) Fail: high het/ low sign and median is not sig
(8) Pass: high het/ low sign and median is sig
(9) Fail: 50+ inst, low sign or high het
(10) Pass: 50+ inst, high sign or low het
```

```{r}

threshold <- 1.4e-7

load("../data/snps_gwas.rdata")
## loads a
gkeep <- filter(a, grepl("mr", data_source.exposure))
gkeeps <- gkeep %>% group_by(id.exposure, trait, exposure) %>% summarise(n=n()) %>% arrange(desc(n)) %>% ungroup %>% filter(!duplicated(trait))
rm(a)
load("../../05_cis-trans-networks/data/outcomes.rdata")
ao <- subset(ao, access %in% c("public", "Public", "immunobase_users"))
table(gkeeps$trait %in% ao$trait)
subset(gkeeps, !trait %in% ao$trait)
category <- subset(ao, trait %in% gkeeps$trait) %>% filter(!is.na(category)) %>% filter(!duplicated(trait)) %>% dplyr::select(trait, category, subcategory)
gkeeps <- merge(gkeeps, category, by="trait")
remove <- c(
	grep("Father", gkeeps$trait),
	grep("Mother", gkeeps$trait),
	grep("Parent", gkeeps$trait),
	grep("survival", gkeeps$trait),
	grep("Difference", gkeeps$trait),
	grep("Subjective", gkeeps$trait),
	grep("Melanoma", gkeeps$trait)
)
remove
gkeeps <- gkeeps[-remove,]
dim(gkeeps)
save(gkeeps, file="../data/gkeeps.rdata")
kable(head(gkeeps))
```

How many traits?

```{r}
nrow(gkeeps)
```

How many disease traits?

```{r}
table(gkeeps$category)
```

Read in tophits and set code to 0. By the end all 0 codes should be removed.

```{r}
load("../results/mrbase_tophits_full.rdata")
res <- subset(res, id.exposure %in% gkeeps$id.exposure)
het <- subset(het, id.exposure %in% gkeeps$id.exposure)
plei <- subset(plei, id.exposure %in% gkeeps$id.exposure)
res$code <- paste(res$id.exposure, res$id.outcome)
het$code <- paste(het$id.exposure, het$id.outcome)
het <- subset(het, method == "Inverse variance weighted")
plei$code <- paste(plei$id.exposure, plei$id.outcome)

codes <- data_frame(code=unique(res$code), decision = 0)
table(codes$decision)
```

### (1) Fail: Non-significant wald ratio

```{r}
temp <- subset(res, nsnp == 1 & pval > threshold)$code
codes$decision[codes$code %in% temp] <- 1
table(codes$decision)
```

### (2) Pass: Significant wald ratio

```{r}
temp <- subset(res, nsnp == 1 & pval <= threshold)$code
codes$decision[codes$code %in% temp] <- 2
table(codes$decision)
```

### (3) Fail: ivw not sig

```{r}
temp <- subset(res, method == "Inverse variance weighted" & pval > threshold)$code
codes$decision[codes$code %in% temp] <- 3
table(codes$decision)
```

### (4) Pass: 2-9 inst, no heterogeneity


```{r}
ivw_followup <- subset(res, method == "Inverse variance weighted" & pval <= threshold & nsnp >= 2 & nsnp < 10)$code
het_followup <- subset(het, code %in% ivw_followup)
temp <- subset(het_followup, Q_pval > 0.05/nrow(het_followup))$code
codes$decision[codes$code %in% temp] <- 4
table(codes$decision)
```

### (5) Fail: heterogeneity

```{r}
temp <- subset(het_followup, Q_pval <= 0.05/nrow(het_followup))$code
codes$decision[codes$code %in% temp] <- 5
table(codes$decision)
```

### (6) Pass: 10-49 inst, no het or high sign

```{r}
ivw_followup <- subset(res, method == "Inverse variance weighted" & pval <= threshold & nsnp >= 10 & nsnp < 50)
het_followup <- subset(het, code %in% ivw_followup$code)
temp1 <- subset(het_followup, Q_pval > 0.05/nrow(het_followup))$code
temp2 <- subset(res, code %in% ivw_followup$code & method == "Sign concordance test")
temp2 <- subset(temp2, pval < 0.05/nrow(temp2))$code
temp3 <- subset(ivw_followup, code %in% c(temp1, temp2))$code

codes$decision[codes$code %in% temp3] <- 6
table(codes$decision)
```

### (7) Fail: high het/ low sign and median is not sig

```{r}
temp4 <- subset(ivw_followup, ! code %in% c(temp1, temp2))
temp5 <- subset(res, code %in% temp4$code & method == "Simple median")
temp6 <- subset(temp5, pval > 0.05/nrow(temp5))$code
codes$decision[codes$code %in% temp6] <- 7
table(codes$decision)
```

### (8) Pass: high het/ low sign and median is sig

```{r}
temp7 <- subset(temp5, pval <= 0.05/nrow(temp5))$code
codes$decision[codes$code %in% temp7] <- 8
table(codes$decision)
```

### (9) Fail: 50+ inst, low sign or high het

```{r}
ivw_followup <- subset(res, method == "Inverse variance weighted" & pval <= threshold & nsnp >= 50)

temp <- subset(res, code %in% ivw_followup$code & method == "Sign concordance test")
temp2 <- subset(het, code %in% ivw_followup$code)

temp3 <- subset(temp, pval < 0.05/nrow(temp))$code
temp4 <- subset(temp2, Q_pval > 0.05/nrow(temp2))$code

temp5 <- subset(ivw_followup, !code %in% c(temp3, temp4))$code
codes$decision[codes$code %in% temp5] <- 9
table(codes$decision)
```

### (10) Pass: 50+ inst, high sign or low het

```{r}
temp6 <- subset(ivw_followup, code %in% c(temp3, temp4))$code
codes$decision[codes$code %in% temp6] <- 10
table(codes$decision)
```

Summarise

```{r}
codes$sig <- codes$decision %in% c(2,4,6,8,10)
codes$msig <- codes$decision %in% c(4,6,8,10)
```

How many are putatively significant

```{r}
table(codes$sig)
```

How many are significant that involve more than one instrument

```{r}
table(codes$msig)
```

Save:

```{r}
save(codes, file="../results/mrbase_sig_codes.rdata")
```

---

Write supplementary table

```{r}
load("../results/mrbase_sig_codes.rdata")
load("../results/mrbase_tophits_full.rdata")
res$code <- paste(res$id.exposure, res$id.outcome)
res <- inner_join(res, codes) %>% filter(method %in% c("Wald ratio", "Inverse variance weighted"))
res$trait <- strsplit(res$exposure, split="\\|") %>% sapply(function(x) x[1]) %>% gsub(" $", "", .)

ressig <- subset(res, sig)
ressig1 <- subset(res, decision == 2)
ressig1 %>% filter(grepl("smok", exposure, ignore.case=TRUE))
ressigm <- subset(res, msig)
ressigm %>% group_by(trait) %>% summarise(n=n())
ressigm %>% group_by(outcome) %>% summarise(n=n())
```

Which single instrument traits are coming up?

```{r}
subset(res, decision == 2) %>% group_by(trait) %>% summarise(n=n()) %>% arrange(desc(n)) %>% kable
```

Which multi-instrument traits?

```{r}
subset(res, msig) %>% group_by(trait) %>% summarise(n=n()) %>% arrange(desc(n)) %>% kable
```

How many traits have at least one effect?

```{r}
subset(res, sig) %>% filter(!duplicated(trait)) %>% nrow
```

How many putative associations (before filtering) but with more than one instrument?

```{r}
sum(res$nsnp > 1 & res$pval < threshold)
```

How many traits?

```{r}
subset(res, nsnp > 1 & pval < threshold) %>% 
filter(!duplicated(trait)) %>% nrow
```

How many traits have at least one effect and more than one instrument, after filtering?

```{r}
subset(res, msig) %>% filter(!duplicated(trait)) %>% nrow
```

How many associations?

```{r}
subset(res, msig) %>% nrow
```

Problem is that sometimes a trait is instrumented by more than one variant but the SNP isn't always available in the analysed CpGs? So some traits appear as having 2 and 1 instruments

```{r}
mtr <- subset(res, nsnp > 1 & pval < threshold) %>% 
filter(!duplicated(trait)) %>% .$trait

str <- subset(res, nsnp == 1 & pval < threshold) %>% 
filter(!duplicated(trait)) %>% .$trait
mtr[mtr %in% str]
```

For simplicity in reporting, do not include these in the single instrument trait list.

```{r}
length(str[!str %in% mtr])
```

How many CpGs involved in the multiple instrument traits?

```{r}
subset(res, msig) %>% filter(!duplicated(outcome)) %>% nrow
```

```{r}
a <- subset(res, nsnp > 1, select=-c(id.exposure, id.outcome, method, chunk, code, sig, exposure))

a <- subset(res, nsnp > 1, select=c(trait, outcome, nsnp, b, se, pval, decision, msig)) %>%
	arrange(desc(msig), pval)
write.csv(a, file="../results/trait-cpg-sig-codes.csv")
```
