---
title: Colocalisation sensitivity analysis
author: Gibran Hemani
---


The `coloc` method may perform at different levels depending on how sparse the SNP data is available to it. In GoDMC the sparseness is non-random - more strongly associated SNPs are more likely to be present. Simulations were conducted to evaluate how this might influence the efficacy of the method.


```{r}
library(dplyr)
library(ggplot2)
load("../data/coloc_results.rdata")
```

There are two methods being used: 
- `fill` = where SNPs are missing fill them in with small effects / high p-values 
- `sparse` = where SNPs are missing only use the ones that are available

When the simulations involve shared causal variants, evaluate how likely the `coloc` method is to detect a shared variant. When the simulations involve distinct causal varients evaluate the false discovery rate of the `coloc` method. Allow the coverage to be complete or incomplete. General picture:

```{r}
group_by(res, coloc, coverage==1, method) %>%
summarise(c=sum(coloc_result==5, na.rm=TRUE)/sum(!is.na(coloc_result)))
```

Plot by coverage

```{r}
res %>%
	group_by(ncausal, coverage, coloc, method) %>%
	summarise(n = sum(!is.na(coloc_result)), pow = sum(coloc_result == 5, na.rm=T) / n) %>%
	ggplot(., aes(x=coverage, y=pow)) +
	geom_point(aes(colour=method)) +
	facet_grid(ncausal ~ coloc)

```